%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   Copyright 2009, Oracle and/or its affiliates.
%   All rights reserved.
%
%
%   Use is subject to license terms.
%
%   This distribution may include materials developed by third parties.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Memory and Memory Operations}
\seclabel{memory-ops}

\note{Check for \emph{initializing write} is not yet supported.}

In this specification,
the term \emph{memory} refers to a set of abstract \emph{locations};
the memory is used to model sharing and mutation.
A location has an associated type,
and \emph{contains} a value of that type
(i.e., the type of the value is a subtype of the type of the location).
Locations can have non-object trait types; by contrast, a value always has an object type.

There are three operations that can be performed on memory:
\emph{allocation}, \emph{read}, and \emph{write}.
Reads and writes are collectively called \emph{memory accesses}.
Intuitively,
a read operation takes a location
and returns the value contained in that location,
and a write operation takes a location and a value of the location's type
and changes the contents of the location
so the location contains the given value.
Accesses need not take place in the order in which they occur in the program text; a detailed account of memory behavior appears in \chapref{memory-model}.

Allocation creates a new location of a given type.
Allocation occurs when a mutable variable is declared, or when a
reference object is constructed.  In the latter case, a new location
is allocated for each field of the object.
Abstractly, locations are never reclaimed;
in practice, memory reclamation is handled by garbage collection.

A freshly allocated location is \emph{uninitialized}.
The type system and memory model in Fortress guarantee that an
\emph{initializing write} is performed to every location if it is ever read,
and that this write occurs before any read of the location.
Any location whose value can be written after initialization is
\emph{mutable}.  Mutable locations consist of
mutable variables and mutable fields of reference objects.
Any location whose value cannot be written after
initialization is \emph{immutable}.
Immutable locations are the non-\KWD{settable} immutable fields of a reference object.
