%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   Copyright 2009 Sun Microsystems, Inc.,
%   4150 Network Circle, Santa Clara, California 95054, U.S.A.
%   All rights reserved.
%
%   U.S. Government Rights - Commercial software.
%   Government users are subject to the Sun Microsystems, Inc. standard
%   license agreement and applicable provisions of the FAR and its supplements.
%
%   Use is subject to license terms.
%
%   This distribution may include materials developed by third parties.
%
%   Sun, Sun Microsystems, the Sun logo and Java are trademarks or registered
%   trademarks of Sun Microsystems, Inc. in the U.S. and other countries.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Dotted Field Accesses}
\seclabel{field-access}

\begin{Grammar}
\emph{Primary} &::=& \emph{Primary}\EXP{.}\emph{Id}\\
\end{Grammar}

An expression consisting of a single subexpression (called the
\emph{receiver expression}), followed by `\txt{.}', followed by an
identifier, not immediately followed by a parenthesis or a white
square bracket, is a
\emph{field access}.
The receiver expression must denote an object
(called the \emph{receiver}) and the field access is evaluated to a call
to a getter mapped from that identifier by the receiver.
The type of the field access is the return type of its getter.
The static type of
the receiver indicates whether a getter mapped from that identifier is
provided by the denoted object.  If a getter is not provided, it is a
static error.
\marginnote{Not exactly: An object (i.e., the actual value at run time)
may provide a getter but if the static type of the receiver expression doesn't
provide the getter, then the getter can't be called in this context.  So just say that
it is a static error if the static type of the receiver doesn't provide the getter,
without making any claim about whether the denoted object provides the getter.}
See \secref{methods} for a discussion of getters.
The evaluation of the receiver must complete normally before the body of
the getter is evaluated.
