%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   Copyright 2009, Oracle and/or its affiliates.
%   All rights reserved.
%
%
%   Use is subject to license terms.
%
%   This distribution may include materials developed by third parties.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Label and Exit}
\seclabel{label-expr}

\begin{Grammar}
\emph{DelimitedExpr} &::=&
\KWD{label} \emph{Id} \emph{BlockElems} \KWD{end} \emph{Id} \\

\emph{FlowExpr}
&::=& \KWD{exit} \option{\emph{Id}} \options{\KWD{with} \emph{Expr}} \\

\end{Grammar}

An expression block may be labeled using a \KWD{label} expression,
which consists of \KWD{label}, an identifier (the block's \emph{name}),
an expression block (its \emph{body}), \KWD{end}, and finally its name again
(it is a static error if the identifier after \KWD{end} is not the name).
The name of a \KWD{label} expression
is in scope in the label namespace
at any point in its body
that is not within a \KWD{spawn} subexpression of the body.
A \KWD{label} expression is evaluated
by evaluating its body.

An \KWD{exit} expression consists of \KWD{exit},
an optional identifier (the \emph{target})
and an optional \KWD{with} clause,
which consists of \KWD{with} followed by an expression.
If a target is specified, it is
a static error if the target is not in scope
in the label namespace at the \KWD{exit} expression.
That is, the target must be the name
of a statically enclosing \KWD{label} expression,
and the \KWD{exit} expression must not be within
a \KWD{spawn} expression that is contained
in the \KWD{label} expression.
If no target is specified, the target is implicitly the name
of the smallest statically enclosing \KWD{label} expression; it is
a static error if there is no such expression.

An \KWD{exit} expression with a \KWD{with} clause
evaluates its \KWD{with} clause expression to yield a \KWD{with} clause value.
The \KWD{exit} expression completes abruptly with the \KWD{with} clause value
(see \secref{eval-completion}).
An \KWD{exit} expression with no \KWD{with} clause
has an implicit \KWD{with} clause whose expression is \EXP{()}.
The type of an \KWD{exit} expression is \TYP{BottomType}.

If the evaluation of the body of a \KWD{label} expression completes normally,
its value is the value of the body.
If the evaluation of the body completes abruptly
with an \KWD{exit} expression whose target
is the name of the \KWD{label} expression,
then the evaluation of the \KWD{label} expression completes normally
and its value is the \KWD{with} clause value of the \KWD{exit} expression.
The type of a \KWD{label} expression
is the union of the type of the last expression of its expression block
and the types of the \KWD{with} clause values of
any \KWD{exit} expressions within the \KWD{label} expression
whose target is the \KWD{label} expression's name.

If one or more \KWD{try} expressions are nested between an
\KWD{exit} expression and the targeted \KWD{label} block, the \KWD{finally}
clauses of these expressions are run in order, from innermost to
outermost, as described in \secref{try-expr}.  If any \KWD{finally}
clause completes abruptly by throwing an exception, the \KWD{exit}
expression fails to exit, the evaluation of the \KWD{label} expression
completes abruptly, and the exception is propagated.

Here is a simple example:
\input{\home/basic/examples/Expr.Label.tex}
The expression ``\EXP{\KWD{exit} I_{95} \KWD{with} \VAR{x32B}}'' completes
abruptly and attempts to transfer control to the end of the
targeted labeled block ``\EXP{\KWD{label} I_{95}}''.
The targeted labeled block completes normally with value \VAR{x32B}.
