%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   Copyright 2009, Oracle and/or its affiliates.
%   All rights reserved.
%
%
%   Use is subject to license terms.
%
%   This distribution may include materials developed by third parties.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Object Expressions}
\seclabel{object-expr}

\note{Property declarations and \KWD{where} clauses are not yet supported.

No naked type variables in \KWD{extends} clause.

The type (tag?) of the {\bf object} resulting from evaluating an object expression is
this anonymous object trait type, but is the static type of the object expression that?

No new functional methods within object expressions.
}

\begin{Grammar}
\emph{DelimitedExpr} &::=&
\KWD{object} \option{\emph{ExtendsWhere}} \option{\emph{GoInAnObject}} \KWD{end} \\

\end{Grammar}

Object expressions denote object values.  Syntactically, they start
with \KWD{object}, followed by an optional \KWD{extends} clause, a
series of field declarations, method declarations,
or property declarations,
and finally \KWD{end}.  The type of an object expression
is an anonymous object expression type that extends the traits listed in the
\KWD{extends} clause of the object expression.  The object expression type does
not include the methods introduced by the object expression (i.e.,
those methods not provided by any supertraits of the object expression).
Each object expression type is associated with a program location;
every evaluation of a given object expression has the same
object expression type.  Two object expressions at different program locations
have different object expression types.


Unlike object declarations (described in \chapref{objects}),
object expressions are not allowed to include
modifiers,
value parameters,
static parameters, or
\KWD{where} clauses (described in \chapref{trait-parameters}).
Object expressions may include free static variables
unlike object declarations, which must
not include any free static variables (i.e., each static variable in
an object declaration must occur
either as a static parameter
or as a \KWD{where}-clause variable).
An object expression is not allowed to declare ``new''
functional methods; it can provide a functional method
only with a name that is already declared by one of its supertraits.
Field initializers and methods may refer to any
variables that are in scope in the context in which the object
expression occurs.  As with object declarations, initializers are evaluated
in textual program order and may refer to previous fields; the object
being constructed might not be referred to in any way.


In an object expression,
any declaration of \KWD{self} (including implicit declarations)
and any declaration which declares a name of a field or dotted method provided
(declared or inherited) by the object expression
in a block enclosing the expression is shadowed
unless the declaration is a method declaration
of a trait extended by the object expression.
In order for an object to be referred to within a nested object expression,
the outer object must be renamed before the object expression
because \EXP{\mathord{\KWD{self}}} declared in the outer object
is shadowed in the object expression:
\input{\home/basic/examples/Expr.Object.a.tex}



For example, the following object expression:
\input{\home/basic/examples/Expr.Object.b.tex}
has a static variable \VAR{T} that is not its static parameter
nor its \KWD{where}-clause variable.



The following example expression evaluates to a new object extending
trait \TYP{Tree}:
\input{\home/basic/examples/Expr.Object.c.tex}
